Meeting the Challenges of Modeling Astrophysical Thermonuclear Explosions: \castro, \maestro, and the \amrex\ Astrophysics Suite

### 2017-10-27 astronum_2017/astronum-amrex.tex
Take a pass through

\address{$^5$Nvidia Corporation}

@@

Among the most significant open questions for SNe Ia is the identity of the
progenitor.  About 20 years ago, the community had
mostly converged upon the single-degenerate scenario---a
Chandrasekhar-mass C/O white dwarf that accretes from its companion,
eventually leading to runaway at the center that burns through the
star (see~\cite{hillebrandtniemeyer2000} for the state of the field at
that time).  Since then, the wealth of observations has indicated that
there is a lot of diversity in SNe Ia, and searches for progenitor
systems have strongly suggested that Chandra-mass white dwarfs cannot
explain most SNe Ia.  Today, merging white dwarfs (the double
degenerate scenario) have perhaps become the most popular model.
Other progenitors, like He burning on the surface of a sub-Chandra
white dwarf, have also seen interest in explaining some of the observed
diversity.  See~\cite{araa-maoz} for a review.

@@

\castro~\cite{castro,castroII,castroIII} is a fully-compressible
radiation hydrodynamics code that supports arbitrary equations of
state, nuclear reaction networks, and Poisson gravity using geometric
multigrid.  The main hydrodynamics scheme in \castro\ is an unsplit
piecewise parabolic method.  The radiation solver in \castro\ uses the
flux-limited diffusion approximation for gray or multigroup radiation.
The integration algorithm on the grid hierarchy is a recursive
procedure in which coarse grids are advanced in time, fine grids are
advanced multiple steps to reach the same time as the coarse grids and
the data at different levels are then synchronized. The
synchronization for self-gravity is similar to the algorithm
introduced by \cite{miniati-colella}.  Recent developments in
\castro\ include a spectral-deferred correction method of coupling
hydrodynamics and reactions, a conservative gravity and rotation
source formulation~\cite{wdmergerI}, and retry mechanism to redo a
step based on criteria evaluated during the
integration.\MarginPar{more new stuff?}

@@

Our latest focus has been on GPU ports of our application codes.  A
small proxy app, \starlord, was created from \castro\ with just the
hydrodynamics and stellar equation of state.  It uses a simple
method-of-lines formulation of the hydrodynamics and advects 13
nuclear species in addition to the hydrodynamics.  To offload work on
GPUs, GPU support was added directly into \amrex\footnote{This support
  is currently on a feature branch in the git repo, awaiting merge
  into {\tt development}.}.  In \amrex, an iterator loops over all of
the boxes at the same level of refinement and passes a data pointer
into a Fortran kernel function where the work is done. For the
simulation state data that resides in each box, we have modified the
memory allocator so that it can use a CUDA allocator (either device or
managed memory).  The domain iterator is configured to handle data
motion to and from the device, so that the compute kernels can operate
on data that is presumed to already be there, and the computation is
decoupled from the memory management. Computation on the data can then
be performed with OpenACC, CUDA Fortran, or (more recently) OpenMP
4.5. We have also built CUDA compute support into \amrex\ so that a
compute kernel can be transparently operated on using CUDA Fortran
without substantially modifying the kernels (and we anticipate using a
similar strategy to use OpenACC and/or OpenMP in the future). This
helps ensure that we can continue to maintain performance portability
in our simulation codes, by decoupling the physics algorithms from the
backend support used to implement them on various compute
architectures.\MarginPar{add CPU time for reference}

Figure~\ref{fig:knl-gpu-scaling} shows the performance of
\starlord\ on the Summitdev platform at OLCF\footnote{Summitdev
  consists of 2 IBM Power8 processors and 4 Nvidia Pascal GPUs per
  node.}.  The highest GPU count (128) corresponds to 32 nodes on
Summitdev.  We see a nearly linear speedup with the number of GPUs,
indicating good weak scaling on this machine.  Efforts are underway to
complete the port of the GPU developments into \castro.  This figure
also shows our performance on the Intel Xeon Phi manycore processors (using
NERSC Cori).  This is for the same problem as the self-gravity test on
Titan.  The Intel Xeon Phi chips in Cori
have 68 cores than can be run with 1, 2, or 4 hardware threads each.
For all these runs, we ran with 4 threads / core (272
threads / chip).

@@

\MarginPar{we should update this to a more modern processor---Don can you run
this on summitdev?}

@@

\ack The work at Stony Brook was supported by DOE/Office of Nuclear
Physics grant DE-FG02-87ER40317 and NSF award AST-1211563.  The work
at LBNL was supported by the DOE Office of Advanced Scientific
Computing Research under Contract No, DE-AC02-05CH11231. An award of
computer time was provided by the Innovative and Novel Computational
Impact on Theory and Experiment (INCITE) program.  This research used
resources of the Oak Ridge Leadership Computing Facility at the Oak
Ridge National Laboratory, which is supported by the Office of Science
of the U.S. Department of Energy under Contract
No.\ DE-AC05-00OR22725.  We appreciate the efforts of the OLCF (and in
particular Fernanda Foertter) in organizing the GPU hackathons.  This
research used resources of the National Energy Research Scientific
Computing Center, which is supported by the Office of Science of the
U.S. Department of Energy under Contract No.\ DE-AC02-05CH11231.
Results in this paper were obtained using the high-performance LIred
computing system at the Institute for Advanced Computational Science
at Stony Brook University, which was obtained through the Empire State
Development grant NYS \#28451.  The GPU work benefited from a grant of
an Titan X Pascal board from Nvidia Corp through the GPU Grant
Program.  Visualization were done using \cite{yt}.  This research has
made use of NASA's Astrophysics Data System Bibliographic Services.

>>>>

\address{$^5$NVIDIA Corporation, 2788 San Tomas Expressway,
  Santa Clara, CA, 95050 USA}

@@

Among the most significant open questions for SNe Ia is the identity of the
progenitor.  About 20 years ago, the community had
mostly converged upon the single-degenerate scenario---a
Chandrasekhar-mass C/O white dwarf that accretes from its companion,
eventually leading to runaway at the center that burns through the
star (see~\cite{hillebrandtniemeyer2000} for the state of the field at
that time).  Since then, a wealth of observations has indicated that
there is a lot of diversity in SNe Ia, and searches for progenitor
systems have strongly suggested that Chandra-mass white dwarfs cannot
explain most SNe Ia.  Today, merging white dwarfs (the double
degenerate scenario) have perhaps become the most popular model.
Other progenitors, like He burning on the surface of a sub-Chandra
white dwarf, have also seen interest in explaining some of the observed
diversity.  See~\cite{araa-maoz} for a review.

@@

\castro~\cite{castro,castroII,castroIII} is a fully-compressible
radiation hydrodynamics code that supports arbitrary equations of
state, nuclear reaction networks, and Poisson gravity using geometric
multigrid.  The main hydrodynamics scheme in \castro\ is an unsplit
piecewise parabolic method.  The radiation solver in \castro\ uses the
flux-limited diffusion approximation for gray or multigroup radiation.
The integration algorithm on the grid hierarchy is a recursive
procedure in which coarse grids are advanced in time, fine grids are
advanced multiple steps to reach the same time as the coarse grids and
the data at different levels are then synchronized. The
synchronization for self-gravity is similar to the algorithm
introduced by \cite{miniati-colella}.  Recent developments in
\castro\ include a spectral-deferred correction method of coupling
hydrodynamics and reactions, a conservative gravity and rotation
source formulation~\cite{wdmergerI}, and a retry mechanism to redo a
step based on criteria evaluated during the
integration.

@@

Our latest focus has been on GPU ports of our application codes.  A
small proxy app, \starlord, was created from \castro\ with just the
hydrodynamics and stellar equation of state.  It uses a simple
method-of-lines formulation of the hydrodynamics and advects 13
nuclear species in addition to the hydrodynamics.  To offload work on
GPUs, GPU support was added directly into \amrex\footnote{This support
  is currently on a feature branch in the git repo, awaiting merge
  into {\tt development}.}.  In \amrex, an iterator loops over all of
the boxes at the same level of refinement and passes a data pointer
into a Fortran kernel function where the work is done. For the
simulation state data that resides in each box, we have modified the
memory allocator so that it can use a CUDA allocator (mainly relying on
managed memory).  The domain iterator is configured to handle data
motion to and from the device, so that the compute kernels can operate
on data that is presumed to already be there, and the computation is
decoupled from the memory management. Computation on the data can then
be performed with OpenACC, CUDA Fortran, or (more recently) OpenMP
4.5. We have also built CUDA compute support into \amrex\ so that a
compute kernel can be transparently operated on using CUDA Fortran
without substantially modifying the kernels (and we anticipate using a
similar strategy to use OpenACC and/or OpenMP in the future). This
helps ensure that we can continue to maintain performance portability
in our simulation codes, by decoupling the physics algorithms from the
backend support used to implement them on various compute
architectures.

Figure~\ref{fig:knl-gpu-scaling} shows the performance of
\starlord\ on the Summitdev platform at OLCF\footnote{Summitdev
  consists of 2 IBM Power8 processors and 4 NVIDIA Pascal GPUs per
  node.}.  The highest GPU count (128) corresponds to 32 nodes on
Summitdev.  We see a nearly linear speedup with the number of GPUs,
indicating good weak scaling on this machine. A single P100 GPU achieves
a performance approximately 2.5 times that of the 20 Power8 cores. Efforts are underway to
complete the port of the GPU developments into \castro.  This figure
also shows our performance on the Intel Xeon Phi manycore processors (using
NERSC Cori).  This is for the same problem as the self-gravity test on
Titan.  The Intel Xeon Phi chips in Cori
have 68 cores than can be run with 1, 2, or 4 hardware threads each.
For all these runs, we ran with 4 threads / core (272
threads / chip).

@@

@@

\ack The work at Stony Brook was supported by DOE/Office of Nuclear
Physics grant DE-FG02-87ER40317 and NSF award AST-1211563.  The work
at LBNL was supported by the DOE Office of Advanced Scientific
Computing Research under Contract No, DE-AC02-05CH11231. An award of
computer time was provided by the Innovative and Novel Computational
Impact on Theory and Experiment (INCITE) program.  This research used
resources of the Oak Ridge Leadership Computing Facility at the Oak
Ridge National Laboratory, which is supported by the Office of Science
of the U.S. Department of Energy under Contract
No.\ DE-AC05-00OR22725.  We appreciate the efforts of the OLCF (and in
particular Fernanda Foertter) in organizing the GPU hackathons.  This
research used resources of the National Energy Research Scientific
Computing Center, which is supported by the Office of Science of the
U.S. Department of Energy under Contract No.\ DE-AC02-05CH11231.
The authors would like to thank Stony Brook Research Computing
and Cyberinfrastructure, and the Institute for Advanced Computational
Science at Stony Brook University for access to the high-performance
LIred and SeaWulf computing systems, the latter of which was made
possible by a \$1.4M National Science Foundation grant (\#1531492).
The GPU work benefited from a grant of an Titan X Pascal board
from NVIDIA Corporation through the GPU Grant Program.
Visualizations were done using \cite{yt}.  This research has
made use of NASA's Astrophysics Data System Bibliographic Services.
